Synthesizing Quotient Lenses

### 2018-03-09 papers/QuotientLenses/main.tex
Final few comments for now
add some changes to eval

\newcommand{\FINISH}[3]{\ifdraft\textcolor{#1}{[#2: #3]}\fi}
\newcommand{\bcp}[1]{\FINISH{dkred}{B}{#1}}
\newcommand{\BCP}[1]{\FINISH{dkred}{B}{\bf #1}}
\newcommand{\afm}[1]{\FINISH{dkgreen}{A}{#1}}
\newcommand{\dpw}[1]{\FINISH{dkblue}{D}{#1}} % Toronto Maple Leafs Blue :-)
\newcommand{\saz}[1]{\FINISH{orange}{SZ}{#1}}
\newcommand{\SAZ}[1]{\FINISH{orange}{SZ}{\bf #1}}
\newcommand{\ksf}[1]{\FINISH{teal}{K}{#1}}
\newcommand{\sam}[1]{\FINISH{dkpurple}{SM}{#1}}

@@

{\em Quotient lenses} loosen many of the restrictions of bijective lenses by
allowing a programmer to define an equivalence relation on the source and
target formats so that the round-tripping lens laws hold modulo these
equivalence rather than modulo equality. Quotient lenses thus make it possible
to align many more data sources than would otherwise be possible, by ignoring
details that a programmer views as inessential. The only language we are aware
of that implements a general class of quotient lenses is {\em
Boomerang}~\cite{quotientlenses}\bcp{Is this the right citation for
Boomerang?  Or should it be \cite{boomerang}?}. While Boomerang's quotient
lenses enable a 
programmer to define a very general class of quotient lenses\bcp{read
  literally, that part of the sentence is nonsense: boomerang's lenses
  (themselves) do not enable the programmer to define lenses.  Of course,
  the reader can probably figure out what is meant, but that's a needless
  tax on their finite mental energy.}, these lenses to have two main
drawbacks.

@@

For the\bcp{why ``the''?  Are we talking about some set of QRE examples
  already?} QRE examples, the $x_i$'s are in $W(c_1)$ and the $y_i$'s are 
in $W(c_2)$, so we can construct ${x'_i} = \canonizer(c_1)(x_i)$ in $K(c_1)$
and ${y'_i} = \canonizer(c_2)(y_i)$ is in $K(c_2)$. To synthesize the
desired quotient lens $q: c_1 \Leftrightarrow c_2$ that is consistent with the
input-output examples $\{(x_1, y_1), \ldots, (x_n, y_n)\}$ it suffices to
synthesize a bijective lens $\ell : K(c_1) \Leftrightarrow K(c_2)$ that is
consistent with the canonized examples $\{({x'_1}, {y'_1}), \ldots, ({x'_n},
{y'_n})\}$ and then apply the canonizers at the outside. Our procedure for QRE
lens synthesis is thus
\begin{enumerate}
  \item
  Compute $K(c_1)$, $K(c_2)$,
  \item
  Compute $\{({x'_1}, {y'_1}), \ldots, ({x'_n}, {y'_n})\}$ where ${x'_i} =
  \canonizer(c_1)(x_i)$ and ${y'_i} = \canonizer(c_2)(y_i)$,
\item Synthesize, using Optician's algorithm, a bijective lens
  $\ell : K(c_1) \Leftrightarrow K(c_2)$ that is consistent with the examples
  $\{({x'_1}, {y'_1}),\allowbreak \ldots,\allowbreak ({x'_n}, {y'_n})\}$
  \item
  Return $q = \kw{rquot}(\kw{lquot}(c_1, \ell), c_2) : c_1 \Leftrightarrow
  c_2$ \bcp{punctuate this enumeration as if it were running text: ``, and''
  at the end of (3) and period here.  And lowercase letters at the
  beginnings of the items.}
\end{enumerate}

\saz{Do we want to state formally the completeness of this algorithm?}\BCP{+1}

@@

In this case, one example is sufficient to uniquely determine the desired
transformation.\SAZ{Is that true?}\BCP{Seems more likely that the desired
  transformation happened to be the first one that the synthesis algorithm
  reached.  Indeed, my intuition is that {\em every} synthesis problem with
  a finite set of examples has infinitely many solutions.  Is that right?}

\SAZ{We might want to say something more precise about the running time or
  expected performance of this algorithm, perhaps to set up the evaluation
  section.  Something along the lines of: ``Although the algorithm is complete
  in theory, in the worst case, it may take an unbounded [???IS THIS
  CORRECT?]\bcp{Surely it's bounded by some exponential formula involving the
    size of the inputs...?} 
  amount time to search for a satisfactory lens.  Nevertheless, this algorithm
  performs well in practice on numerous examples, as we show in the next
  section.}

@@

We have implemented QREs and the quotient lens synthesis algorithm described
above as an extension to the Boomerang interpreter~\cite{boomerang,quotientlenses}. Thanks to this integration, QREs and quotient lens
synthesis tasks can now be expressed alongside existing Boomerang
terms\bcp{what's the difference between a task and a term (and how do each
  of them relate to ``lenses'' just below)?}, and we
can combine the two kinds of lenses seamlessly in the same application.
Intuition suggests that the increased expressiveness of QREs should make many
lens programming tasks easier, but we would like more quantitative understanding
of that claim.  We thus conducted several experiments that aim to provide
answers to the following questions:

\begin{enumerate}
  \item Does synthesizing quotient lenses from QREs permit an easier development
  process than existing approaches?  \bcp{Needlessly vague: What existing approaches?  Writing
  them by hand?}

  \item Does the increase in expressivity gained from synthesizing from QREs
  permit synthesis of more useful lenses than existing approaches.  \bcp{Is
    this really an interesting empirical question?  Seems obvious, from the
    arguments we made in Section 2 and the intro.}

  \item Does our tool synthesize lenses fast enough to be used as part of a
  standard development process.  \bcp{Should there be some question marks
    around here.}
\end{enumerate}

@@

\item The remaining eight benchmarks were derived from datasets found at 
  data.gov. These datasets include census
  population data, wage comparison data, demographics statistics data and
  crime index data. We chose these datasets since they were stored in XML, CSV
  and RDF formats which are more easily specified using regular expressions than
  plain text or PDF documents.  \bcp{Why did we include these eight new
    ones?  Are they significantly different (more complex, exercising
    different features of the synthesis algorithm or QRE syntax, ...)?}
\end{enumerate}

@@

\begin{figure}[t]
\includegraphics{generated-graphs/asts.eps}
\caption{Count of benchmark programs definable using a given AST count. We find
that it takes far fewer AST nodes to define benchmark lenses using QRE
synthesis than with Optician without QRE synthesis or without synthesis.\saz{Why
is the bar for ``All QRE Benchmarks'' at 14? Shouldn't it be at 47 = 39+8? Are
we not testing all QRE benchmarks?}}
\label{fig:asts}
\end{figure}

To get a sense of how much the new expressiveness that combining QRE lenses with
synthesis saves in terms of programmer effort, we compared the amount of code
that would need to be written for each of the lens benchmarks.  We use number of
nodes in the abstract syntax tree as a measure of code size, and calculated
three quantities for each lens synthesis benchmark task.  \saz{Say something
  about why there are only 14 benchmarks?}
%
\begin{itemize}
  \item[\QRESize{}] the number of AST nodes in the QRE
  specifications (including examples)\saz{CHECK THIS}
  \item[\CanonizerAndSpecSize{}]  the number of
  AST nodes in $W(q_1)$ summed with the number of AST nodes in $\canonizer(q_1)$
  \item[\LensAndSpecSize{}] the number of AST nodes
  in $W(q_1)$ summed with the number of AST nodes in the synthesized QRE lens.
\end{itemize}

@@

The value of \CanonizerAndSpecSize{} estimates the burden for writing our
benchmarks within the Optician system, without QREs but still using
synthesis. (This is only an approximation, as both $W(q_1)$ and $\canonizer(q_1)$ are
automatically generated from the QRE, and it is possible that a human-written
version might be smaller.)

The value of \LensAndSpecSize{} estimates the burden in writing our benchmarks
in Boomerang, using neither QREs nor synthesis. (This is also an
approximation, as $W(q_1)$ and $\canonizer(q_1)$ and the bijective lens are all
automatically generated.)

Figure~\ref{fig:asts} shows the number of benchmarks that can be expressed using
at most a given number of AST nodes. For a given benchmark, we found \QRESize{}
to be about half as large as \CanonizerAndSpecSize, and about a third as large
as \LensAndSpecSize{}.\BCP{These are really the critical conclusions, and
  they are hard to see, the way the graph is drawn.  I wonder whether we
  could instead present with the individual examples on the X axis (sorted
  by number of AST nodes), with a bar above each one containing three
  regions, colored differently, for LS, CS-LS, and QC-CS...} This suggests
that introducing QREs saves programmers 
significant effort compared to both Optician and basic Boomerang.

@@

We found that 10 of Optician's 39 benchmarks required modifications to be
bijective. We found that 4 of Optician's 39 benchmarks required modifications to
be in bijective correspondence when in canonical form.\saz{Is this somehow where
  the 14 = 10+4 points in Fig. 10 come from?}\saz{Do these 10 and 4 intersect?}
Furthermore, we found that all eight of the data.gov examples required no
modifications to be in bijective correspondence when in canonical form. While we
did not put the data.gov examples into a form suitable for pre-QRE Optician, we
found that only four of the eight data.gov examples did not require
canonization, so making them suitable for pre-QRE Optician would have required 
changes to of four of the data.gov formats.  Here again, we see that QREs add
significant expressive power.

@@

Also notable is one of the data.gov examples, which converts demographic
statistics by zip code from xml form into rdf form. This took longer because it
was an exceptionally large format, converting over 40 fields.\saz{We need to
  explain more detail here.  I guess this is about the last benchmark in Fig. 11
  (b), which takes a long time?}  We added in additional fields to make this
format suitable for Optician, which took over 15 seconds synthesize the modified
lens.\saz{This needs more context, explanation.}

>>>>

\newcommand{\FINISH}[3]{\ifdraft\textcolor{#1}{[#2: #3]}\fi}
\newcommand{\bcp}[1]{\FINISH{dkred}{B}{#1}}
\newcommand{\BCP}[1]{\FINISH{dkred}{B}{\bf #1}}
\newcommand{\afm}[1]{\FINISH{dkgreen}{A}{#1}}
\newcommand{\dpw}[1]{\FINISH{dkblue}{D}{#1}} % Toronto Maple Leafs Blue :-)
\newcommand{\saz}[1]{\FINISH{orange}{SZ}{#1}}
\newcommand{\ksf}[1]{\FINISH{teal}{K}{#1}}
\newcommand{\sam}[1]{\FINISH{dkpurple}{SM}{#1}}

@@

{\em Quotient lenses} loosen many of the restrictions of bijective lenses by
allowing a programmer to define an equivalence relation on the source and
target formats so that the round-tripping lens laws hold modulo these
equivalence rather than modulo equality. Quotient lenses thus make it possible
to align many more data sources than would otherwise be possible, by ignoring
details that a programmer views as inessential. The only language we are aware
of that implements a general class of quotient lenses is {\em
Boomerang}~\cite{quotientlenses}. While Boomerang's quotient lenses enable a
programmer to define a very general class of quotient lenses, these lenses to
have two main drawbacks.

@@

For the\bcp{why ``the''?  Are we talking about some set of QRE examples
  already?} QRE examples, the $x_i$'s are in $W(c_1)$ and the $y_i$'s are 
in $W(c_2)$, so we can construct ${x'_i} = \canonizer(c_1)(x_i)$ in $K(c_1)$
and ${y'_i} = \canonizer(c_2)(y_i)$ is in $K(c_2)$. To synthesize the
desired quotient lens $q: c_1 \Leftrightarrow c_2$ that is consistent with the
input-output examples $\{(x_1, y_1), \ldots, (x_n, y_n)\}$ it suffices to
synthesize a bijective lens $\ell : K(c_1) \Leftrightarrow K(c_2)$ that is
consistent with the canonized examples $\{({x'_1}, {y'_1}), \ldots, ({x'_n},
{y'_n})\}$ and then apply the canonizers at the outside. Our procedure for QRE
lens synthesis is thus
\begin{enumerate}
  \item
  Compute $K(c_1)$, $K(c_2)$,
  \item
  Compute $\{({x'_1}, {y'_1}), \ldots, ({x'_n}, {y'_n})\}$ where ${x'_i} =
  \canonizer(c_1)(x_i)$ and ${y'_i} = \canonizer(c_2)(y_i)$,
\item Synthesize, using Optician's algorithm, a bijective lens
  $\ell : K(c_1) \Leftrightarrow K(c_2)$ that is consistent with the examples
  $\{({x'_1}, {y'_1}),\allowbreak \ldots,\allowbreak ({x'_n}, {y'_n})\}$
  \item
  Return $q = \kw{rquot}(\kw{lquot}(c_1, \ell), c_2) : c_1 \Leftrightarrow
  c_2$ \bcp{punctuate this enumeration as if it were running text: ``, and''
  at the end of (3) and period here.}
\end{enumerate}

\saz{Do we want to state formally the completeness of this algorithm?}

@@

In this case, one example is sufficient to uniquely determine the desired
transformation.\saz{Is that true?}

\saz{We might want to say something more precise about the running time or
  expected performance of this algorithm, perhaps to set up the evaluation
  section.  Something along the lines of: ``Although the algorithm is complete
  in theory, in the worst case, it may take an unbounded [???IS THIS CORRECT?]
  amount time to search for a satisfactory lens.  Nevertheless, this algorithm
  performs well in practice on numerous examples, as we show in the next
  section.}

@@

We have implemented QREs and the quotient lens synthesis algorithm described
above as part of Boomerang. Thanks to this integration, QREs and quotient lens
synthesis tasks can now be expressed alongside existing Boomerang terms, and we
can combine the two kinds of lenses seamlessly in the same application.
Intuition suggests that the increased expressiveness of QREs should make many
lens programming tasks easier, but we would like more quantitative understanding
of that claim.  We thus conducted several experiments that aim to provide
answers to the following questions:

\begin{enumerate}
  \item Does synthesizing quotient lenses from QREs permit an easier development
  process than existing approaches?

  \item Does the increase in expressivity gained from synthesizing from QREs
  permit synthesis of more useful lenses than existing approaches.

  \item Does our tool synthesize lenses fast enough to be used as part of a
  standard development process.
\end{enumerate}

@@

\item The remaining eight benchmarks were derived from
  data.gov.\footnote{\url{http://data.gov}} These datasets include census
  population data, wage comparison data, demographics statistics data and
  crime index data. We chose these datasets since they were stored in XML, CSV
  and RDF formats which are more easily specified using regular expressions than
  plain text or PDF documents.
\end{enumerate}

@@

\begin{figure}[t]
\includegraphics{generated-graphs/asts.eps}
\caption{Count of benchmark programs definable using a given AST count. We find
that it takes far fewer AST nodes to define benchmark lenses using QRE
synthesis than with Optician without QRE synthesis or without synthesis.}
\label{fig:asts}
\end{figure}

To get a sense of how much the new expressiveness that combining QRE lenses with
synthesis saves in terms of programmer effort, we compared the amount of code
that would need to be written for each of the lens benchmarks. We use number of
nodes in the abstract syntax tree as a measure of code size, and calculated
three quantities for each lens synthesis benchmark task. We only compared the
sizes on benchmarks where the canonization functions are nontrivial.
%
\begin{itemize}
  \item[\QRESize{}] the number of AST nodes in the QRE
  specifications (including examples)\saz{CHECK THIS}
  \item[\CanonizerAndSpecSize{}]  the number of
  AST nodes in $W(q_1)$ summed with the number of AST nodes in $\canonizer(q_1)$
  \item[\LensAndSpecSize{}] the number of AST nodes
  in $W(q_1)$ summed with the number of AST nodes in the synthesized QRE lens.
\end{itemize}

@@

The value of \CanonizerAndSpecSize{} estimates the burden for writing our
benchmarks within the Optician system, without QREs but still using
synthesis. This is only an approximation, as both $W(q_1)$ and $\canonizer(q_1)$ are
automatically generated from the QRE, and it is possible that a human-written
version might be smaller.

The value of \LensAndSpecSize{} estimates the burden in writing our benchmarks
in Boomerang, using neither QREs nor synthesis. This value is also an
approximation, as $W(q_1)$ and $\canonizer(q_1)$ and the bijective lens are all
automatically generated.

Figure~\ref{fig:asts} shows the number of benchmarks that can be expressed using
at most a given number of AST nodes. For a given benchmark, we found \QRESize{}
to be about half as large as \CanonizerAndSpecSize, and about a third as large
as \LensAndSpecSize{}. This demonstrates that introducing QREs saves programmers
significant effort compared to both Optician and basic Boomerang.

@@

We found that 10 of Optician's 39 benchmarks required modifications to be
bijective. Of these We found that 4 of Optician's 39 benchmarks required
modifications to be in bijective correspondence when in canonical form.
Furthermore, we found that all eight of the data.gov examples required no
modifications to be in bijective correspondence when in canonical form. While we
did not put the data.gov examples into a form suitable for pre-QRE Optician, we
found that only four of the eight data.gov examples did not require
canonization, so making them suitable for pre-QRE Optician would have required
changes to of four of the data.gov formats. Here again, we see that QREs add
significant expressive power.

@@

Also notable is one of the data.gov examples, which converts demographic
statistics by zip code from xml form into rdf form. To validate this performance
was not due to our overhead, we modified this format to be bijective, so it
could be input into Optician. Optician took over 15 seconds synthesize this
modified lens. We believe this conversion took longer because the input formats
were exceptionally large, requiring conversions between over 40 fields.
